---
title: "R Self-Study Guide: Data Structures, Functions, and Visualization"
author: "Samuel Irungu Kigotho"
date: "2025-10-22"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 

This R script is a tutorial on fundamental R concepts: vectors, matrices, arrays, data frames, lists, functions, loops, apply-family functions, and data visualization with ggplot2

# Vectors
- A vector is simply a list of items that are of the same type.
- *c()* combines values into a vector (the simplest R data structure).
- *val* is a numeric vector with 6 elements.


```{r }
val = c(1, 2, 3, 4, 5, 6)
val
class(val)
```
- Mixing numbers and text converts all elements into character type.
- R uses coercion — all elements in a vector must be the same type.

```{r}
x = c(1, 2.0, 3.0, 4, 5, "Hello", "OK")
x
class(x)

```
- Naming elements inside a vector — useful for labeling values.
- Access by label: temperature["morning"].

```{r}
temperature = c(morning = 20, before_noon = 23, after_noon = 25, evening = 22, night = 18)
temperature["morning"]
```
## Basic Operations with Vectors
- Vectors of equal length can be added element-wise.
```{r}
jan_price = c(10, 20, 30)
june_price = c(20, 25, 33)
mar_price = jan_price +june_price
mar_price
```
## Character Vectors
```{r}
items = c("potato", "rice", "oil")
class(items)

```

# Matrix
A matrix is a two dimensional data set with columns and rows.
A column is a vertical representation of data, while a row is a horizontal representation of data.
A matrix can be created with the *matrix()* function. Specify the *nrow* and *ncol* parameters to get the amount of rows and columns:
  - Lets Combines all prices into one matrix.
  - *nrow* = 3 → matrix will have 3 rows.
  - *byrow* = TRUE fills the matrix row by row instead of column by column.

```{r}
all_prices = matrix(c(jan_price, mar_price, june_price), nrow = 3)
all_prices
class(all_prices)
all_prices2 = matrix(c(jan_price, mar_price, june_price), nrow = 3, byrow = TRUE)
all_prices2
```

# Array
An array is a data structure that can hold multi-dimensional data. In R, the array is objects that can hold two or more than two-dimensional data
  - Lets combine 6 vectors into a 3×3×2 array:
   - 3 rows → items
   - 3 columns → months
   - 2 layers → years (2018 and 2017)
```{r}
# Create six vectors
jan_2018 = c(10, 11, 20)
mar_2018 = c(20, 22, 25)
june_2018 = c(30, 33, 33)
jan_2017 = c(10, 10, 17)
mar_2017 = c(18, 23, 21)
june_2017 = c(25, 31, 35)

combined = array(c(jan_2018, mar_2018, june_2018, jan_2017, mar_2017, june_2017), dim = c(3, 3, 2))
combined

```



## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
